CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems

### 2014-12-09 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
More updates to the partitioning sections of the paper.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4524 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

%   - Impact of # of cutlines
\subsection{Impact of Number of Dice}
\label{num_dice_impact}
TODO: Javeed
To examine the impact of the number of die used to construct an interposer-based FPGA, we varied the number of cuts from 1 to 3 (which varies the number of die from 2 to 4).  In these experiments, the \textit{delay increase} was kept constant at $1ns$.

As shown in Figure \ref{fig:cuts_minW}, the number of cuts does have an impact on the minimum channel width, but not a very large or constant one. At 80\% of wires cut the experiments with 2 and 3 cuts had the minimum channel width only 8\% and 10\%, respectively, greater than the scenario with only 1 cut. At lower values of \textit{\% wires cut} the difference was even smaller.

Figure \ref{fig:cuts_crit} shows that the number of dice in an interposer-based FPGA impacts circuit speed significantly, as the critical path delay rises significantly as systems have more cuts. Recall that a monolithic FPGA has a geometric average critical path delay of $23.7ns$ for our test architecture. With \textit{number of cuts} $= 1$ the critical path delay increases by 3 to 4 ns, while with \textit{number of cuts} $= 2$ the critical path delay increases by 5 to 6 ns. With \textit{number of cuts} $= 3$, the critical path delay is typically 7 ns higher than that of a monolithic FPGA.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{cuts_minW.eps}
\caption{Minimum channel width vs. \textit{\% wires cut} for 1, 2 and 3 cuts and $1ns$ of \textit{delay increase}.}
\label{fig:cuts_minW}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{cuts_crit_path.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 1, 2 and 3 cuts and $1ns$ of \textit{delay increase}.}
\label{fig:cuts_crit}
\end{figure}

%   - Impact of aspect ratio
\subsection{Impact of Aspect Ratio}
\label{aspect_ratio_impact}
TODO: Javeed

>>>>

%   - Impact of # of cutlines
\subsection{Impact of Number of Dice}
\label{num_dice_impact}
To examine the impact of the number of dice used to construct an interposer-based FPGA, we measured the minimum channel width for an FPGA device with four dice, varying the fraction of wires cut. We used the best CAD flow from Section \ref{sec:partitioner_stage_results} and the best architecture parameters from this section (TODO JAVEED -- spell them out?). The results, shown in Figure \ref{fig:wires_cut_4part}, show the same trend as Figure \ref{fig:wires_cut}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{wires_cut_4part.eps}
\caption{Impact of partitioning-based CAD flow on routability for a range of architectures with 4 dice. TODO JAVEED -- LU32 90\% result is missing}
\label{fig:wires_cut_4part}
\end{figure}

TODO JAVEED -- show critical path delay? (it's uninteresting)

%   - Impact of aspect ratio
\subsection{Impact of Aspect Ratio}
\label{aspect_ratio_impact}
TODO: Javeed
We investigated the implications of the device aspect ratio on routability for the four-partition case. In particular, we compared a device with a square interposer (four rectangular dice) to a device with a rectangular interposer (four square dice.) We used the best CAD flow from Section \ref{sec:partitioner_stage_results} and the best architecture parameters from this section. The results are shown in Table \ref{table:squares}.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|r|r|}
\hline
 & Square Interposer & Square Dice \\ \hline \hline
bgm & 102 & 306 \\ \hline
stereovision0 & 86 & 256 \\ \hline
stereovision1 & 166 & 256 \\ \hline
LU8PEEng & 156 & 126 \\ \hline
mkDelayWorker32B & 96 & 194 \\ \hline
mcml & 196 & 496 \\ \hline
LU32PEEng & 256 & 264 \\ \hline
geomean & 140.7481367 & 251.805728 \\ \hline
\end{tabular}
\caption{Minimum channel width for square interposer and square dice configurations.}
\label{table:squares}
\end{table}

The square interposer configuration performs significantly better than the square dice configuration.

TODO JAVEED -- Vaughn commented that we should talk about which is easier to manufacture. I think the square dice device would actually be easier to manufacture, because the more complex dice will fit well on a round wafer. Isn't it fine for the interposer to be non-square?

### 2014-12-09 doc/journal/fpgapaperfinal/FPGAPaper/partitioning.tex
More updates to the partitioning sections of the paper.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4524 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\subsubsection{Partitioner Stage}
As shown in Figure \ref{fig:partitioning_cad_flow}, we explore the effect of circuit partitioning at two different stages of the CAD flow. The first possibility is to partition the netlist of primitive elements (LUTs, FFs, RAM slices, etc.) before packing into the larger function blocks such as logic blocks and (32 kb) RAM blocks. In this flow the partitioning constraints must be respected by the packer; it must not pack two primitives assigned to different partitions (dice) into the same function block. We updated the VPR clustering algorithm to respect this new legality constraint.

@@

\begin{enumerate}
  \item Partitioning the primitive netlist to guide the packer, leaving the placer unconstrained.
  \item Partitioning the primitive netlist to guide the packer, and constraining the placer based on constraints derived from partitioning, as described above.
  \item Leaving the packer unconstrained, and partitioning the function block (clustered) netlist to guide the placer.
  \item Leaving both the packer and the placer unconstrained. This is the base flow that does not use partitioning, as described in Section \ref{cadSection}.
\end{enumerate}

@@

\subsubsection{Hypergraph to Graph Transformation}
We transformed the circuit netlist hypergraph to a graph in the following way:
\begin{enumerate}
\item For each hyperedge of the circuit netlist hypergraph, we generated a graph (the \emph{per-net subgraph}).
\item We combined all of the per-net subgraphs, summing the edge weights for edges appearing in more than one per-net subgraph, to generate the total netlist graph.
\end{enumerate}

@@

To validate the choice of hyperedge cutsize as a proxy for circuit routability, we ran the partitioning CAD flow for each graph topology and edge weight scheme, across several circuits. We use an unbalance of $5\%$, split the clustered netlist into two partitions (constraining only the placer), cut $80\%$ of the wires crossing the interposer, and impose a $1ns$ delay penalty for wires crossing the interposer. To accommodate unbalance in the placement engine, we increase the size of the grid of complex blocks that make up the FPGA device. Relative to the minimum device size required for placement (without partitioning constraints), we add $10\%$ to the complex block grid width and $10\%$ to the grid height. Figure ?? shows the geometric mean of the minimum channel width required for a successful route, for each of the graph topology and edge weight combinations.

>>>>

\subsubsection{Partitioner Stage}\label{sec:partitioner_stage}
As shown in Figure \ref{fig:partitioning_cad_flow}, we explore the effect of circuit partitioning at two different stages of the CAD flow. The first possibility is to partition the netlist of primitive elements (LUTs, FFs, RAM slices, etc.) before packing into the larger function blocks such as logic blocks and (32 kb) RAM blocks. In this flow the partitioning constraints must be respected by the packer; it must not pack two primitives assigned to different partitions (dice) into the same function block. We updated the VPR clustering algorithm to respect this new legality constraint.

@@

\begin{enumerate}
  \item Leaving both the packer and the placer unconstrained. This is the base flow that does not use partitioning, as described in Section \ref{cadSection}.
  \item Partitioning the primitive netlist to guide the packer, leaving the placer unconstrained.
  \item Leaving the packer unconstrained, and partitioning the function block (clustered) netlist to guide the placer.
  \item Partitioning the primitive netlist to guide the packer, and constraining the placer based on constraints derived from partitioning, as described above.
\end{enumerate}

@@

\subsubsection{Hypergraph to Graph Transformation}\label{sec:hypergraph_to_graph}
We transformed the circuit netlist hypergraph to a graph in the following way:
\begin{enumerate}
\item For each hyperedge of the circuit netlist hypergraph, we generated a graph (the \emph{per-net subgraph}).
\item We combined all of the per-net subgraphs, summing the edge weights for edges appearing in more than one per-net subgraph, to generate the total netlist graph.
\end{enumerate}

@@

To validate the choice of hyperedge cutsize as a proxy for circuit routability, we ran the partitioning CAD flow for each graph topology and edge weight scheme, across several circuits. We use an unbalance of $5\%$, split the clustered netlist into two partitions (constraining only the placer), cut $80\%$ of the wires crossing the interposer, and impose a $1ns$ delay penalty for wires crossing the interposer. To accommodate unbalance in the placement engine, we increase the size of the grid of complex blocks that make up the FPGA device. Relative to the minimum device size required for placement (without partitioning constraints), we add $10\%$ to the complex block grid width and $10\%$ to the grid height. Figure \ref{fig:graph_topology_mcw} shows the geometric mean of the minimum channel width required for a successful route, for each of the graph topology and edge weight combinations.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{graph_topology_mcw.eps}
\caption{Minimum channel width achieved by VPR with different hyperedge models and weightings. n is the number of vertices on a hyperedge.}
\label{fig:graph_topology_mcw}
\end{figure}

The star topology with 1/n edge weights achieves the best minimum channel width, confirming that hyperedge cutsize is a good proxy for routability. Consequently, we use this scheme in all future results in this paper. Unlike routability, we found that the post-routing critical path delay was not strongly impacted by the hypergraph conversion scheme used.

\subsubsection{Partitioner Stage Results}\label{sec:partitioner_stage_results}
We compare the performance of the four CAD flow variations described in Section \ref{sec:partitioner_stage} on the 8 largest VTR benchmarks. We again use an unbalance of $5\%$ and split the clustered netlist into two partitions (constraining the packer and/or the placer, as required by the flow under test). The interposer parameters (\% wires cut = $80\%$ and delay increase = 1 ns) and bloat factor ($10\%$ in each dimension) are the same as in Section \ref{sec:hypergraph_to_graph}. Figure \ref{fig:flows_mcw} shows the minimum routable channel width for each partitioning CAD flow variation.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{vpr_flows_mcw.eps}
\caption{Minimum channel width achieved by VPR with the different partitioning flows.}
\label{fig:flows_mcw}
\end{figure}

On average, the best minimum channel width was obtained when constraining both packing and placement. However, constraining packing alone performed significantly better than constraining placement alone. This shows that the VPR packer strongly benefits from partitioning constraints when targeting an interposer-based architecture, likely because it forbids the packing of primitives that are not naturally related into one function block. Interestingly, other recent work targeting conventional FPGAs has shown that using a partitioner to guide packing benefits routability [Wenyi Feng, et al, “Rent’s Rule Based Packing for Routability Optimization,” FPGA 2014 ]. On the other hand, The VPR placement algorithm with the enhancements described in Section 4 sees only a modest routability benefit from the partitioning constraints.

For the same set of circuits, we computed the critical path delay at a channel width equal to 1.3x the minimum. The results are shown in Figure \ref{fig:flows_crit_path}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{vpr_flows_crit_path.eps}
\caption{Critical path delay achieved by VPR with the different partitioning flows.}
\label{fig:flows_crit_path}
\end{figure}

On average, the critical path delay did not vary significantly between the different partitioning CAD flows, but it appears that partitioning before packing may slightly improve circuit speed, while enforcing these partitioning constraints during placement may slightly reduce circuit speed.

Considering both routability and delay, the best CAD flow constrained both packing and placement with the partitioning result and this is the flow we use in the remainder of this paper.

\subsubsection{\% wires cut}
To see how well the CAD flow works for a range of interposer architectures, we varied the fraction of wires cut and measured the average minimum channel width over the 8 largest VTR circuits. As shown in Figure \ref{fig:wires_cut}, this new flow is a significant improvement over that of [] for FPGAs where more than half the wires are cut at interposer boundaries. It is slightly worse than the traditionally FPGA CAD flow even for an architecture with 0\% of the wires cut, which is a monolithic FPGA, presumably because we have restricted the placer's freedom.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{wires_cut.eps}
\caption{Impact of partitioning-based CAD flow on routability for a range of architectures.}
\label{fig:wires_cut}
\end{figure}

\subsubsection{Bloating of the Clustered Netlist due to Partitioning Constraints}
The additional constraints on the packer imposed by partitioning may result in a less dense packing. Figure \ref{fig:packing_bloat} compares the number of complex blocks in the partitioning-constrained and unconstrained clustered netlists.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{packing_bloat.eps}
\caption{Packing bloat.}
\label{fig:packing_bloat}
\end{figure}

On average, the packing-constrained netlist contained $3.2\%$ more complex blocks than the unconstrained netlist. The LU8PEEng and LU32PEEng circuits were the most affected by this packing bloat. In those circuits, we found that the packing bloat was due almost entirely to poor packing of memory blocks; an interesting area for future work would be to augment the partitioner to understand that some RAM primitives are best kept together in one partition.
